Ocular biomarker profiling after complement factor I gene therapy in geographic atrophy secondary to age-related macular degeneration

Version 1

Abstract

Abstract

Objective

Complement biomarker analysis in ocular fluid samples from subjects with geographic atrophy (GA) in a Phase I/II clinical trial of subretinal AAV2 complement factor I (CFI; FI) gene therapy, PPY988 (formerly GT005), to understand target pharmacokinetics/pharmacodynamics. Clinical findings were subsequently utilized to investigate the therapeutic dose in an in vitro complement activation assay.

Design, setting and participants

Biomarker data were evaluated from 28 subjects in FOCUS, a Phase I/II clinical trial evaluating the safety and efficacy of three ascending doses of PPY988.

Main outcomes and measures

Vitreous humor (VH), and aqueous humor (AH) from subjects before surgery and at serial timepoints (week 5 or 12, 36, 96) were evaluated for changes in levels of intact complement factors I, B and H (FI, FB, FH) components C3, C4, and C1q and breakdown products (Ba, C3a, C3b/iC3b, C4b) using validated assays and OLINK® proteomics.

A modified in vitro assay of complement activation modelling VH complement concentrations was used to compare PPY988 potency to the approved intravitreal C3 inhibitor pegcetacoplan (Apellis) and complement Factor H (FH).

Results

An average 2-fold increase in VH FI was observed post-treatment at week 36 and week 96. This correlated with a marked post-treatment reduction in VH concentration of the FB breakdown product Ba and Ba:FB ratio, but minimal changes in C3a and C3b/iC3b levels. Variable concordance in complement biomarker levels in VH versus AH suggest AH is not a reliable proxy for VH for complement activation. During the experimental comparison of doses, a 2-fold increase of FI achieved in the vitreous had only a minor effect on the complement amplification loop in vitro, indicating limited impact [IC50: 1229nM]. Pegcetacoplan completely blocks C3a generation at concentrations much lower than the estimated trough level for monthly intravitreal injections [IC50: 2nM]. Supplementation with FH in the assay revealed similar potency to pegcetacoplan [IC50: 6nM].

Conclusions and relevance

PPY988 subretinal gene therapy may not have provided sufficient FI protein to meaningfully modulate complement activation to slow GA growth. Reviewing VH biomarkers is important for understanding target expression, pathway engagement, and determining optimal dose, thereby informing future clinical development.

Introduction

Age-related macular degeneration (AMD) is a leading cause of blindness, particularly in developed countries. In the United States (US) alone, there are approximately 18 million individuals aged 40 and above with early-stage AMD, and 1.5 million with late-stage AMD, which includes geographic atrophy (GA) and/or neovascular AMD 1. GA is a condition characterized by the progressive loss of photoreceptors, retinal pigment epithelium (RPE), and underlying choriocapillaris, resulting in a central scotoma 2. Reduced vision can significantly impact activities of daily living and quality of life 3,4. There have been two recent Food and Drug Administration (FDA)-approved therapies that inhibit complement activity, with a modest slowing of GA growth 5,6. However, they require life-long monthly or bimonthly intravitreal (IVT) injections and have so far exhibited limited functional efficacy that present challenges for long term compliance 7. Therefore, a one-time gene therapy that provides long-lasting efficacy equivalent to or better than repeated IVT therapies is a promising avenue for exploration.

One such gene therapy, called PPY998 (formerly GT005), is a subretinal gene therapy that supplements complement factor I (CFI; FI) protein in the retinal layers where atrophy occurs 8. A Phase II study (HORIZON, NCT04566445) was initiated, however a decision to stop the PPY988 program was taken by Novartis based on a recommendation from the independent Data Monitoring Committee following an overall benefit-risk assessment of available data from the program studies. [Press release: Novartis AG; September 11th 2023,: https://www.novartis.com/news/gt005-ppy988-development-program-geographic-atrophy]. This raises important questions about whether CFI was sufficiently expressed after subretinal gene therapy, whether there was evidence of modulation of intraocular complement activity, or if this modulation was sufficient to potentially alter the disease course. Such information is crucial to the retina community, as there are many companies developing complement modulating therapies for AMD 9. Sharing the learnings from the PPY998 program would improve the risk-benefit analysis for these programs and help understand if new approaches differentiate from PPY998, justifying further clinical study of complement targets for GA.

The complement system, specifically the alternative pathway (AP), plays an important role in the development of AMD, and increasing local FI concentration could be a promising approach to regulate ocular complement activity. The AP is a part of the complement system that constantly surveys for foreign cells and debris and requires tight regulation to prevent inflammation and self-cell damage 10. Regulatory proteins like factor H (FH), decay-accelerating factor (DAF; CD55), membrane cofactor protein (MCP; CD46), complement receptor 1 (CR1; CD35), and FI help control the AP, with FI being the only regulatory enzyme 11. FI, expressed from the CFI gene, cleaves C3b by forming a trimolecular complex with cofactors FH for C3b, and MCP and CR1 for C3b and C4b 12–18. Studies have shown that carriers of rare variants (RV) in CFI have increased risk of developing advanced AMD and more rapid disease progression 19–23. CFI RV can result in low levels of serum FI (Type I variants) or normal levels with functional impairment (Type II variants) 24,25. Research has also suggested that doubling serum FI levels may help control complement overactivation driven by other AMD risk genetic factors mapping to complement genes 26. Based on these findings, clinical trials were conducted to explore the potential of sub-retinal delivery of AAV2-CFI as a gene therapy for treating dry AMD 8.

The FOCUS study is a Phase I/II open-label study evaluating the safety, and dose response of three doses of PPY988 AAV2-CFI subretinal gene therapy in GA. To assess transgene pharmacokinetics and pharmacodynamics, vitreous humor (VH) and aqueous humor (AH) samples were collected at baseline (BL) and four time points up to 96 weeks after dosing and were evaluated for exploratory complement biomarkers. This report shares findings regarding change from baseline in FI and complement protein levels and activation, over time after vitrectomy and PPY988 administration. Levels of FI achieved in VH from FOCUS subjects were then used in an in vitro simulated human vitreous complement activation assay, to gain insights on the therapeutic potencies of FI, C3 inhibitor pegcetacoplan (Syfovre, Apellis), and FH. We also offer critical learnings for other AMD complement gene therapy programs and a guide for future analyses of complement biomarker data in ocular samples.

Methods

Ethics

The FOCUS study was conducted in accordance with the tenets of the Declaration of Helsinki and all other local regulations, in accordance with the International Conference on Harmonization E6 Guidelines for Good Clinical Practice and with applicable local, state, and federal laws. Institutional review board or ethics committee approval was obtained, and the study was registered at ClinicalTrials.gov (reference: NCT03846193) and EudraCT (reference: 2017-003712-39). All subjects provided written informed consent to participate, and an independent data monitoring committee provided ongoing oversight.

Study design

FOCUS is an open-label first in human Phase I/II multicentre clinical trial to evaluate the safety, and dose response of PPY988 (AAV2-CFI) 8, administered as a single subretinal injection in subjects with GA secondary to AMD, followed for up to 240 weeks (5 years) post-dosing. Subjects were administered one of three doses to the study eye under local anaesthesia by a qualified vitreoretinal surgeon: 2E10 viral genome (vg; low), 5eE10 vg (medium) and 2E11 vg (high). At screening, a saliva sample was taken from subjects and extracted DNA was analysed for CFI RV using targeted next generation sequencing 21. Genotyping results were not required at screening if already available through participation in a previous Gyroscope sponsored study. The term ‘rare’ was defined as variants that changed the CFI coding sequence, had ≤1% minor allele frequency in GnomAD non-Finnish Europeans (V2.1.1, accessed http://gnomad.broadinstitute.org), and were previously associated with FI haploinsufficiency 11. A detailed description of study design, route of administration, dose, number of subjects, biomarker matrices, sampling timing and methodology are provided in Supplementary Methods and Supplementary Table 1.

Complement protein and total protein quantification

AH and VH samples of subjects were taken on the treated eye immediately prior to subretinal surgery at baseline (BL) and at 5 or 12 weeks, 36, and 96 weeks post-treatment, as described in detail in Supplementary Table 1. Samples were rapidly frozen at −80 °C and transported frozen for analysis. Prior to analysis, samples were thawed and subjected to centrifugation at 17,000g for 10 minutes at 4°C.Samples were kept on ice during sample preparation.

Complement factors C2, C4b, C5, C5a, factor D, mannose binding lectin (MBL), FI, C1q, C3b/iC3b, C3, FH, FB and C4 were quantified in AH and VH using the MILLIPLEX Human Complement Panel 1 and 2 (cat. no. HCMP1MAG-19K and HCMP2MAG-19K, EMD Millipore Corporation, Billerica, MA) following extensive quality control and validation of calibration curves. Plates were read using the Luminex MagPix and xPONENT software (Luminex Corporation, Austin, TX). C5a was omitted from further analysis as measurements fell under the lower limit for quantification.

Complement factors C3a and Ba were quantified in AH and VH using MicroVue EIA kits (cat. no. A032 and A034, QuidelOrtho Corporation, San Diego, CA) following manufacturer instructions. Plates were read using a Tecan Sunrise absorbance microplate reader using the Magellan Tracker data analysis software (Tecan Trading AG, Switzerland).

Each assay was validated and considered fit for purpose. Sample processing and analyses were conducted in accordance with Good Clinical Laboratory Practice, the analytes or measurements that did not meet acceptance criteria (coefficient of variation of 20%, or 30% when results are below the lower limit of quantification but above limit of detection) were excluded from the final analysis. The number of successfully sampled subjects for each biomarker at each time-point is reported in Supplementary Table 2.

Total protein was quantified in AH and VH using the Quant-iT protein assay kit (cat. no. Q33210, Invitrogen, Thermo Fisher Scientific, Waltham, MA) as per the manufacturer’s instructions. Plates were read using a Varioskan LUX fluorescent microplate reader using the SkanIt data acquisition software (Thermo Fisher Scientific, Waltham, MA).

In vitro modelling the ocular complement system and potency assay

The assay was performed using reagents from the Wieslab AP assay kit (cat. no. COMPLAP330, Svar Life Science AB, Sweden) and purified human serum complement proteins (Complement Technology Inc., Texas USA. All preparation steps were performed on ice. Serial dilutions of the test analytes FI, FH and pegcetacoplan were separately prepared and added to a mix of the complement components C3, FB, FH, FI and Properdin simulating the final concentrations measured in the VH baseline of FOCUS subjects (CFI RV negative population). The simulated vitreous complement mix was transferred on the lipopolysaccharide (LPS) coated Wieslab plate on ice. FD was separately added to the mix, the plate was immediately put at 37°C and incubated for 1 hour. At the end of the incubation, the solution was transferred into a dilution plate and kept on ice, EDTA was added to each well to obtain a final 20mM concentration. Ba and C3a generated in the assay were measured by using the MicroVUE EIA kits (cat. no. A032 and A034, QuidelOrtho Corporation, San Diego, CA) following the manufacturer instructions. The ELISA plates were scanned using the Varioskan plate reader and analysed using the SkanIT software (Thermo Fisher Scientific, Waltham, MA). The in vitro FI potency comparability assay is described in Supplementary Methods. The trough level of pegcetacoplan in the VH after 27 days was calculated by halving the treatment dose of 15mg/eye 6 times (pegcetacoplan half-life 4.5 days), the number obtained was converted to µg/mL using an estimated volume of VH of 4 mL 27.

Statistics

All subject sample biomarker analyses were performed in R version 4.1.0. The analysis set consisted of all subjects who received subretinal delivery of PPY998 from cohorts 1 to 4 and had at least one non-missing biomarker assessment performed at BL and either week 36 or week 96. Given sparsity of data at week 5 and week 12, the data were combined, and the average value was used in instances where subjects had non-missing data from both visits. Concentrations values above the upper limit of quantification (ULOQ) was set to ULOQ and values below the lower limit of detection (LOD) were set to LOD. Please see Supplementary Tables 3-4 for limits.

Significance testing for the change in protein level from BL to week 36 or week 96 were performed on the differences of log-transformed data with the non-parametric Wilcoxon rank-sum test. Means and standard error bars are shown in the respective figures. Correlations between VH and AH sampling, as well as between FI and other proteins, were summarized with the Spearman’s Rank correlation coefficient. As analyses were exploratory in nature and hypothesis-generating, no adjustments for multiple testing were performed.

For the in vitro VH modelling assay, to calculate IC50s, 4-parameter logistic curves were generated using GraphPad Prism version 10.0.0 for Windows, GraphPad Software, Boston, Massachusetts USA, www.graphpad.com.

Results

Biomarker detection in FOCUS subjects that received PPY988

In FOCUS Part 1 and 2 (cohorts 1 to 4), 31 subjects received one of three doses of PPY988 via the standard subretinal delivery approach (Supplementary Table 1). A total of 97% of the ocular samples expected under the protocol were obtained prior to early termination of the study which triggered cessation of further ocular fluid sampling. A total of 88 VH, 63 AH and 45 plasma samples from 30 subjects were obtained at various timepoints including BL, and 5 or 12, 36 and 96 weeks post-treatment (Supplementary Methods and Supplementary Tables 1 and 2).

In FOCUS, PPY988 was well tolerated with no serious ocular AEs, some dose-related RPE changes in the bleb area were evident and there was no clinically significant vector-related inflammation. A full account of the FOCUS study safety and clinical data, as well as full Phase II PPY988 safety, efficacy and biomarker data will be reported separately following study completion.

Vitrectomy for subretinal AAV-CFI gene therapy procedure results in sustained loss of protein from the VH

To test impact of total vitrectomy performed during transvitreal subretinal delivery of PPY988 (Figure 1A and B) influencing global proteomic matrix changes and biomarker analyses/interpretation, total protein (TP) measurements were obtained at BL and each sampling timepoint (according to sample availability). In VH, a ∼30% reduction in TP from baseline was observed at either week 5 or 12 and week 36 (NS, P=0.162), with a significant >50% reduction identified at week 96 (P=0.033), while there were minimal changes in AH TP at week 5/12, 36 or 96 (Figure 1C). Because VH TP was reduced post-operatively, we performed TP normalisation to compare the results with non-normalised data interpretation for VH but not AH.

Surgical delivery of PPY988 AAV2-CFI.

(A) PPY988 vector design. FI protein is secreted by cells targeted by AAV2 serotype and containing the cDNA sequence for wild-type CFI (NM_000204.4) driven by a CAG promoter 8. (B) Subretinal delivery of PPY988. (C) Vitrectomy during SR delivery of PPY988 results in loss of total protein (TP) in the vitreous humor. TP concentration (y-axis, log2 µg/µL) was measured at baseline and 5 or 12, 36, and 96 weeks post-surgery (y-axis). Mean +/- SE is shown by blue bars and lines whilst all subjects are shown by black lines for both vitreous (left) and aqueous (right) humor. A significant reduction in TP was observed at week 96 in vitreous humor (*P=0.033). Significance is defined as *P<0.05. BL = Baseline; WK= week.

AAV2-CFI delivery results in a sustained increase in FI protein level in the VH and AH

Transgene expression was measured in VH and AH samples at weeks 5 or 12, 36, and 96 after treatment with PPY988. On average, we identified a 2-fold increase in VH FI at week 36 compared to baseline, with non-TP normalized baseline concentrations of 500 ng/mL compared to 1000 ng/mL at week 36 (non-normalised P<0.001; TP normalised increase ∼2-fold (P=0.002)) (Figure 2A, 2B). The increase in FI concentration was sustained, with a ∼2.5-fold increase in FI observed at week 96. (mean ∼1250 ng/mL) (non-normalised P<0.001; TP normalised increase ∼3-fold (P=0.002)). CFI RV (n=5) and non-CFI RV (‘broad’) (n=19) genotype groups demonstrate comparable magnitudes of absolute FI elevation post-treatment, with the overall CFI RV cohort FI starting concentration ∼50% reduced (NS). Elevated FI expression in AH samples at week 36 (P<0.001) were sustained out to week 96 (P<0.001) (Figure 2C, 2D). Lastly, there was a variable response to the three PPY988 doses in both VH and AH with no significant dose effect observed (Figure 2C, 2D).

Temporal expression of factor I (FI) in vitreous humor (VH) and aqueous humor (AH) after PPY988 delivery.

(A,B) FI levels (y-axis, ng/mL) in VH were significantly increased at week 36 and 96 post-surgery (x-axis) both when non-normalised (A) (P=0.001 for both timepoints) and total protein-normalised (B) (P=0.002 at both timepoints). Mean +/- SE is shown by bars for all (black lines) broad (blue lines) and CFI RV (green lines) groups. (C,D) FI level (y-axis, ng/mL) for each subject in the cohort is tracked by week (x-axis) and separated by dose (2E10 vg: orange lines; 5E10 vg: green lines; 2E11 vg: blue lines) in VH (C) and AH (with significant increase in AH FI at week 36 and 96 (both P<0.001)) (D), where mean +/- SE shown by black bars. Significance is defined as *P<0.05; **P<0.01, ***P<0.001. N numbers are shown above X-axis. BL = Baseline; WK= week.

Increased FI expression results in a reduction in Ba and Ba:FB in the VH

To test for FI target engagement at the level of formation of the alternative pathway C3 convertase, we measured Ba and FB concentrations at each sampling week after subretinal delivery of PPY988. A reduction in Ba in the VH is indicative of a reduction in C3 convertase formation in the ocular environment. We observed that Ba levels were reduced by ∼2-fold compared to baseline at week 36 and 2.5-fold at week 96 in non-normalised data (week 36: P=0.004; week 96: P=0.003) (Figure 3A) and TP normalised data (week 36: P=0.002; week 96: P=0.038) (Figure 3B). To add confidence to this read-out by accounting for fluctuations in endogenous FB biosynthesis, ratios of intact complement proteins and their breakdown products were assessed as in previous studies 28–30. Ba:FB ratio changes post-treatment reflected the raw data. When reviewing biomarker levels from both non-normalized and TP normalized datasets, a significant reduction in Ba:FB was observed at week 36 (both P<0.001) and at week 96 (non-TP normalized (P=0.001); TP normalized (P=0.033)) (Figure 3C, 3D).

Modulation of Ba and Ba:FB by PPY988 delivery.

(A,B) Ba concentrations (Y-axis, ng/mL) in vitreous humor (VH) samples by week post-surgery (X-axis). Mean +/- SE is shown by bars for all (black lines) broad (blue lines) and CFI RV (green lines) groups. Significant reductions in Ba were observed at week 36 for both non-normalised (P=0.004) (A) and total protein (TP) normalised (P=002) (B) data and week 96 for non-normalised data (P=0.003) and TP normalised (P=038). (C,D) Ratios of Ba:FB concentration (y-axis, ng/mL) were calculated for each subject at each sampling post-surgery week (x-axis). Mean +/- SE is shown by bars for all (black lines) broad (blue lines) and CFI RV (green lines) groups. Significant reductions were identified at week 36 for both non-normalised (P<0.001) (C) and TP normalised (P<0.001) (D) Ba:FB ratios, with a reduction at week 96 for non-normalized (P=0.001) and TP normalized (P=0.033) ratios. Significance is defined as *P<0.05; **P<0.01, ***P<0.001. N numbers are shown above X-axis. BL = Baseline; WK= week.

Increased FI expression results in modest alterations in VH and AH complement biomarker profiles

To investigate the effect of PPY988 delivery on complement AP and classical pathway (CP) activity, a panel of biomarkers including FI, C3, FB, FH, C1q, Ba, C3a, C3b/iC3b, C4 and C4b were measured in the VH and AH (in addition to plasma) at BL, week 5 or 12, week 36 and week 96. The complete set of concentrations at BL are reported in the Supplementary Table 3 (VH), Supplementary Table 4 (AH), and Supplementary Table 5 (plasma). In addition, Ba:FB, C3a:C3 and C3b/iC3b:C3 ratios were calculated at each time-point. Baseline values for each biomarker with descriptive statistics are reported in Supplementary Table 3 (VH) and Supplementary Table 4 (AH), with baseline and post-treatment pairwise correlations between FI and biomarkers presented in Supplementary Figure 1 (VH) and Supplementary Figure 2 (AH). Fold changes in complement biomarker concentration from baseline are presented for each biomarker or ratio at each post-surgery sampling week in VH non-normalised data (Figure 4A), VH TP normalised data (Figure 4B) and AH non-normalised data (Figure 4C). In the non-normalised VH dataset, we observed reductions in Ba and Ba:FB, with a reduction in C1q possibly indicative of reduced ocular immune cell (e.g. microglial) activation 31 (Figure 4A). We detected no modulation of VH C3a and C4b. Although levels of C3b/iC3b per unit of C3 fall substantially at week 96 in both non-normalized (P=0.002) and TP-normalized (P=0.012) analyses (Supplementary Figure 3), pairwise correlation analyses reveal that no biomarkers or ratios were negatively correlated with increasing FI level at week 36 or 96 post-treatment (Supplementary Figures 1 and 2). Minimal reduction of CP activity was indicated by C4b:C4 ratio analysis, however, this was not observed when normalizing for TP (Supplementary Figure 4). VH FB and FH concentrations do not fluctuate substantially from BL values after treatment with PPY988. Overall, TP normalisation did not substantially alter the interpretation of VH biomarkers with broadly similar trends observed (Figure 4B).

Modulation of complement AP and CP biomarkers by PPY988 delivery.

Fold-changes (y-axis) in complement biomarker [FI, C3, FH, FB, C1q, Ba, C3a, C3b/iC3b, C4b, Ba:FB, C3a:C3, C3b/iC3b:C3] concentrations from baseline at each post-surgery week (VH: week 5/12, 36 and 96; AH: week 36 and 96; X-axis) are shown in VH with no normalisation (A), VH with total protein normalisation (B), and AH with no normalisation. Mean is shown by thick black lines, +/- SE is shown by bars. Individual subject changes are shown by grey lines. Yellow boxes highlight disparate trends in VH vs AH biomarkers. BL = Baseline; WK= week.

AH sampling is not reflective of VH complement activation profile

Complement biomarker profiling in AH demonstrated increasing concentrations of a majority of biomarkers (Figure 4C), including for Ba, C1q, and C3a (at week 96) which was discordant with VH sampling data. Whether these components (C1q) and breakdown products (Ba, C3a) are generated locally, systemically, or diffusely from the retina or VH is unclear. A similar trend to VH was observed when considering the increase in AH C3 and associated fall in C3a:C3 and C3b/iC3b:C3 ratios at week 96 in particular. However, in the case of AH, the increase in C3 was also observed at week 36, suggesting compartmentalization of response of the AH vs the VH. In line with these data, pairwise correlations between VH and AH levels of Ba, C3a and C3b/iC3b post-treatment are modest and variable (Supplementary Figure 5), with only FI consistently well correlated between the two ocular compartments.

Proteomic analysis on VH using Olink provides supportive evidence for target expression and modest modulation of the complement system

Olink proteomic analysis of the VH samples identified a 2-fold median increase in week 36 FI levels relative to baseline (Supplementary Methods, Supplementary Figure 6a). Further review of differentially expressed proteins revealed trends for elevation of additional complement components including factor B, properdin and C2 (Supplementary Figure 6b, Supplementary Table 6).

VH modelling of complement activation reveals inferior potency of FI vs pegcetacoplan

By modelling the components of the VH based on median measurements of C3, FB, FH, FI (broad cohort measurements) and properdin (all measurements) at baseline in the FOCUS study, we compared the potency of FI, FH and the C3 inhibitor pegcetacoplan for their effectiveness in controlling the generation of C3a on an LPS-coated plate in a simulated ocular environment context (Figure 5). In this assay, C3 is directly activated by O-antigens on LPS thus triggering activation to drive the AP amplification loop. A wide range of FI, FH and pegcetacoplan concentrations were titrated through an artificial VH matrix so that comparisons could be made to clinically relevant levels of both FI and pegcetacoplan. Serum purified FI was revealed to be of similar potency to FI expressed from Good Manufacturing Practice (GMP) manufactured PPY988 (Supplementary Figure 7). The trough level of pegcetacoplan at 27 days was estimated using injection dose and half-life information available on the SyfovreTM prescribing information datasheet. No diffusion data in human was available so this was not counted in the trough calculation. The concentration obtained was similar to that reported in the pegcetacoplan rabbit pharmacokinetics (PK) study and PK/PD prediction study published by Apellis 32,33. Pegcetacoplan vitreal trough level was compared to the 1st and 3rd quartile increase (at week 36 vs BL) in FI associated with PPY988 delivery. Pegcetacoplan exhibited greater potency than FI, with complete inhibition of the AP amplification loop and C3a generation at predicted trough level (∼59 μg/mL), whilst FI was observed to have minimal effect at levels of expression associated with PPY988 after 36 weeks. Overall curves generated by logistic regression revealed an IC50 of 1229 nM for FI, whilst FH [IC50: 6 nM] was similar in potency to pegcetacoplan [IC50: 2 nM]. The estimated vitreous trough level of pegcetacoplan was approximately 100-fold greater than that required for complete inhibition of the AP in this assay. On the other hand, FI was observed to have minimal effect on AP activation at the levels of PPY988 measured in VH after 36 weeks of treatment.

In vitro modelling of vitreous humor complement activation inhibition by FI vs pegcetacoplan and additional complement regulators.

FI (orange), FH (purple), and pegcetacoplan (blue) were titrated (x-axis, ng/mL) through artificial vitreous complement matrix on LPS-coated WeislabTM plates and C3a generation was measured by ELISA (y-axis, ng/mL). C3a is a marker of convertase activity in the complement activation cascade, the potency of complement inhibitors or regulators is reflected in lower amount of C3a generated in the assay when testing comparable concentrations. Mean is shown with +/-SD represented by bars and 4-parameter fit logistic regression curves are shown by lines. The 1st-3rd quartile of increased FI expression from baseline after 36 weeks of PPY988 delivery is shown by orange area. The calculated trough concentration of pegcetacoplan in eyes 27 days after IVT injection is shown by the blue vertical line.

Discussion

Gene therapy has been suggested as a promising treatment approach for slowing macular degeneration in GA secondary to AMD. This manuscript describes a comprehensive examination of the pharmacokinetic / pharmacodynamic properties of complement biomarkers in the context of the PPY988 GA subretinal gene therapy program and provides technical considerations for analyzing biomarker data from treated subjects. Despite the early termination of the clinical development program, the PPY988 study achieved notable milestones in the field, including the implementation of subretinal delivery of gene therapy for GA and the use of serial vitreous sampling. Herein, we present for the first time an analysis of VH complement biomarkers at baseline and after surgical delivery of a subretinal gene therapy. Baseline levels were similar to a previous study of VH complement 34 with potentially confounding loss of total protein also observed in our study, highlighting the need to analyze data in the context of changing total protein levels post-vitrectomy. Our observations reveal consistent elevation of VH FI levels over a 96-week period, accompanied by reduced detection of the FB activation product Ba and ratios of Ba:FB, consistent with sustained modulation of complement AP activity in the vitreous humor compartment. These data confirm that AAV gene therapy has a consistent and lasting effect. Furthermore, although unpowered due to low N, the effects of FI gene therapy on complement breakdown product biomarker profiles (Ba (Figure 2A) and C3b/iC3b (Supplementary Figure 3A) appear to be comparable in both CFI RV positive and negative sub-populations, which is counter to the original hypothesis of a more dramatic effect in a sub-population with lower baseline FI levels 19,20. However, due to small sample size this finding should be interpreted with caution. Although the impact on C3 breakdown products was minimal at week 36, it became significant at week 96. Elevation of FI and modulation of the broader complement system was also observed in Olink proteomic profiling of VH samples (Supplementary Figure 6), supporting the complement biomarker findings in the validated panel. Elevated levels of FI were detected in both AH and VH. PPY988 efflux from the sub-retinal injection may have transduced cells in the anterior chamber of the eye resulting in an increase in AH FI concentration. However, the correlation of other complement biomarkers between VH and AH was inconsistent, in particular, Ba and C1q exhibited disparate post-treatment trends (Figure 4), indicating that AH alone may not be a reliable biomarker matrix for assessing the activity of complement biomarkers at the target tissues. This cautionary advice should be considered in context of previous reports, suggesting AH could be a broadly reliable proxy for VH 35, albeit that only full length complement proteins were studied therein. To ensure quality control, protein level normalization was added to the analysis comparing biomarker levels before and after vitrectomy. Interestingly, the overall conclusions remained consistent regardless of protein adjustment, which strengthens confidence in the study’s findings. After determining the futility of the Phase II study (HORIZON, NCT04566445), an in vitro assay using clinically relevant levels of complement proteins was conducted to investigate a possible biomarker-related explanation. The results showed that pegcetacoplan had higher potency than FI in reducing C3 convertase activity in vitro suggesting that limited in-vitro potency of FI may have contributed to the lack of clinical success at Phase II. Although the translatability of the VH modelling assay to GA remains to be demonstrated, FI’s lower potency suggests that further increasing FI concentration on its own by gene therapy may not achieve clinically meaningful levels of C3 inhibition. It should however be noted that intravitreal pegcetacoplan will be at its highest concentration in the mid-vitreous whereas FI will most likely be at the highest level in the subretinal space where it is being produced. Hence directly comparing vitreous levels of each compound may underestimate the activity of FI compared with pegcetacoplan at the level of the subretinal space.

The lack of efficacy in the PPY988 Phase II study does not detract from the viability of complement activation as an effective target for intervention in GA, given the therapeutic effect observed for complement inhibitors pegcetacoplan and avacincaptad pegol (Izervay; Iveric Bio) 5,6, and this study’s observation that pegcetacoplan achieves a much greater potency than FI in vitro. FI had a disproportionally larger effect upon Ba compared to C3a and C3b/iC3b, which may be due to the lability of C3a, differences in diffusion capabilities of the molecules from the retina, and possible contribution of C3 concentration from the choriocapillaris as well as biosynthesis by local immune cells. In addition, we hypothesize FI’s capacity to regulate the production of Ba without causing significant changes in the levels of C3b and C3a in the fluid phase could be contributed to by regulation of AP tickover, resulting from binding and activation of FB on water-hydrolysed C3 (C3H2O.) followed by release of Ba10. Similarly, the intravitreal administration of a neutralizing factor D Fab, which was found to be ineffective in Phase III trials for GA, was also able to reduce ocular Ba levels 28. However, it did not have a significant impact on the generation of VH C3 breakdown products. In this study, by week 96, there was an observed rise in the levels of full-length C3 and a corresponding decrease in the ratios of breakdown products to full length. While the specific implications of this change are uncertain, it potentially suggests a reduction in C3 consumption, which aligns with the known activity of FI, but this may be confounded with upregulated biosynthesis of C3 given its acute phase nature 36. C3 breakdown products only partially diffuse due to cell-surface adherence. Based on this understanding, we hypothesize that changes in C3 breakdown products may only become detectable when a certain threshold of modulation in tissue complement activity is reached. Overall, the data suggests that fully functional FI is being secreted into the VH, but the regulatory effects on the level of Ba may be representative of convertase formation in the VH and not the macula retina/RPE nor the choroid. To validate this hypothesis, one approach would be to conduct vitreal sampling using an effective drug targeting C3 for GA in a larger cohort. Further investigations are needed to gain a better understanding of these possibilities.

The study has several limitations that should be considered when interpreting the results. Firstly, the study did not include a control group, which means that the observed changes in complement biomarkers may not be solely attributed to transgene expression. Other factors, such as the impact of subretinal gene therapy administration triggering local inflammation or the effects of total vitrectomy on the protein environment, could also contribute to these changes. During the study, no significant safety concerns or adverse events were reported aside from dose-responsive incidents of RPE changes in the area of the subretinal injection. However, increased levels of antibodies to the vector and transgene were observed following treatment 37. It is important to note that these antibody levels were transient and resolved shortly after treatment, indicating that the observed antibody responses were not persistent or associated with any significant clinical consequences. It is worth noting that the subjects included in this early clinical trial had more advanced disease compared to the typical clinical trial GA population which may impact the generalizability of the findings to the broader GA population. Additionally, the study had a relatively small sample size, which increases the risk of bias. Furthermore, the biomarker conclusions drawn from the study subjects may not be representative of the subjects treated in the subsequent Phase II study. Therefore, caution should be exercised when generalizing the findings to larger populations. Despite convincing and robust genetic links to AMD reported for Type I CFI variants 19–22, this study observed no difference in biomarker response when comparing CFI RV carriers to the broad population. Interestingly, emerging data have revealed more modest ORs for CFI RVs in progression from early to late AMD 23 and macular thinning was observed in UK Biobank eyes of carriers <60 years old 38. Moreover, the long-term effects of complement therapy on the progression of GA are still uncertain. Another limitation to consider is the translatability of the in vitro VH modelling assay to the disease processes occurring in GA, and the relevance of the assay to cell-surface complement activation occurring in the area of GA. The LPS-coated surface used for the VH modelling assay would most likely result in more acute complement activation than in the GA retina. Finally, this is a relative potency assay and does not take into consideration the biodistribution of different modalities and routes of administration. Despite these limitations, the conclusions drawn from this study are compelling and hold significant implications for other gene therapy programs. The introduction of the innovative in vitro complement activation assay serves as a valuable benchmark for gene therapy programs.

The superior potency of FH compared to FI highlights the importance of conducting potency comparisons using in vitro assays that reflect physiologically relevant levels of complement proteins to guide decision-making. This finding opens promising avenues for further investigation, with potential for engineering of more active RCA protein truncates and chimeras including CR1, MCP, DAF and FH, FHRs. To advance gene therapy programs, future research should focus on building upon the pharmacokinetic / pharmacodynamic findings. It is crucial to establish tissue concentrations and estimates of the target required for complement modulation before moving into clinical stages. Pre-clinical assays that can be effectively translated to GA remain elusive. Overall, the role of complement in GA and the potential benefits of a safe one-time gene therapy in reducing AMD progression suggest the need for further clinical development of complement gene therapy.

eLife Assessment

This important work advances our understanding of factors influencing efficacy assessments and biomarker viability for complement-directed gene therapy against age-related macular degeneration. The data presented is convincing and offers insights and teachings for the design of gene therapy and complement-targeted therapeutics in the eye and more broadly for future ocular biomarker studies.

Reviewer #1 (Public Review):

Summary:

This study analyzed biomarker data from 28 subjects with geographic atrophy (GA) in a Phase I/II clinical trial of PPY988, a subretinal AAV2 complement factor I (CFI) gene therapy, to evaluate pharmacokinetics and pharmacodynamics. Post-treatment, a 2-fold increase in the vitreous humor (VH) FI was observed, correlating with a reduction in FB breakdown product Ba but minimal changes in other complement factors. The aqueous humor (AH) was found to be an unreliable proxy for VH in assessing complement activation. In vitro assays showed that the increase in FI had a minor effect on the complement amplification loop compared to the more potent C3 inhibitor pegcetacoplan. These findings suggest that PPY988 may not provide enough FI protein to effectively modulate complement activation and slow GA progression, highlighting the need for a thorough biomarker review to determine optimal dosing in future studies.

Strengths:

This manuscript provides critical data on the efficacy of gene therapy for the eye, specifically introducing complement FI expression. It presents the results from a halted clinical trial, making sharing this data essential for understanding the outcomes of this gene therapy approach. The findings offer valuable insights and lessons for future gene therapy attempts in similar contexts.

Weaknesses:

No particular weaknesses. The study was carefully performed and limitations are discussed.

I have just some concerns about the methodology used. The authors use the MILLIPLEX assays, which allow for multiplexed detection of complement proteins and they mention extensive validation. How are the measurements with this assay correlating with gold standard methods? Is the specificity and the expected normal ranges preserved with this assay? This also stands for the Olink assay. Some of the proteins are measured by both assay and/or by standard ELISA. How do these measurements correlate?

Reviewer #2 (Public Review):

Summary:

The results presented demonstrate that AAV2-CFI gene therapy delivers long-term and marginally higher FI protein in vitreous humor that results in a concomitant reduction in the FB activation product Ba. However, the lack of clinical efficacy in the phase I/II study, possibly due to lower in vitro potency when compared to currently approved pegcetacoplan, raises important considerations for the utility of this therapeutic approach. Despite the early termination of the PPY988 clinical development program, the study achieved significant milestones, including the implementation of subretinal gene therapy delivery in older adults, complement biomarker comparison between serial vitreous humor and aqueous humor samples and vitreous humor proteomic assessment via Olink.

Strengths:

Long-term augmentation of FI protein in vitreous humor over 96 weeks and reduction of FB breakdown product Ba in vitreous humor suggests modulation of the complement system. Developed a novel in vitro assay suggesting FI's ability to reduce C3 convertase activity is weaker than pegcetacoplan and FH and may suggest a higher dose of FI will be required for clinical efficacy. Warn of the poor correlation between vitreous humor and aqueous humor biomarkers and suggest aqueous humor may not be a reliable proxy for vitreous humor with regard to complement activation/inhibition studies.

Weaknesses:

The vitrectomy required for the subretinal route of administration causes a long-term loss of total protein and may influence the interpretation of complement biomarker results even with normalization. The modified in vitro assay of complement activation suggests a several hundred-fold increase in FI protein is required to significantly affect C3a levels. Interestingly, the in vitro assay demonstrates 100% inhibition of C3a with pegcetacoplan and FH therapeutics, but only a 50% reduction with FI even at the highest concentrations tested. This observation suggests FI may not be rate-limiting for negative complement regulation under the in vitro conditions tested and potentially in the eye. It is unclear if pharmacokinetic and pharmacodynamic properties in aqueous humor and vitreous humor compartments are reliable predictors of FI level/activity after subretinal delivery AAV2-CFI gene therapy.

Reviewer #3 (Public Review):

Summary:

The manuscript by Hallam et al describes the analysis of various biomarkers in patients undergoing complement factor I supplementation treatment (PPY988 gene therapy) as part of the FOCUS Phase I/II clinical trial. The authors used validated methods (multiplexed assays and OLINK proteomics) for measuring multiple soluble complement proteins in the aqueous humour (AH) and vitreous humour (VH) of 28 patients over a series of time points, up to and including 96 weeks. Based on biomarker comparisons, the levels of FI synthesised by PPY988 were believed to be insufficient to achieve the desired level of complement inhibition. Subsequent comparative experiments showed that PPY988-delivered FI was much less efficacious than Pegceptacoplan (FDA-approved complement inhibitor under the name SYFORVE) when tested in an artificial VH matrix.

Strengths:

The manuscript is well written with data clearly presented and appropriate statistics used for the analysis itself. It's great to see data from real clinical samples that can help support future studies and therapeutic design. The identification that complement biomarker levels present in the AH do not represent the levels found in the VH is an important finding for the field, given the number of complement-targeting therapies in development and the desperate need for good biomarkers for target engagement. This study also provides a wealth of baseline complement protein measurements in both human AH and VH (and companion measurements in plasma) that will prove useful for future studies.

Weaknesses:

Perhaps the conclusions drawn regarding the lack of observed efficacy are not fully justified. The authors focus on the hypothesis that not enough FI was synthesised in these patients receiving the PPY988 gene therapy, suggesting a delivery/transduction/expression issue. But beyond rare CFI genetic variants, most genetic associations with AMD imply that it is a FI-cofactor disease. A hypothesis supported by the authors' own experiments when they supplement their artificial VH matrix with FH and achieve a significantly greater breakdown of C3b than achieved with PPY988 treatment alone. Justification around doubling FI levels driving complement turnover refers to studies conducted in blood, which has an entirely different complement protein profile than VH. In Supplemental Table 5 we see there is approx. 10-fold more FH than FI (533ug/ml vs 50ug/ml respectively) so increasing FI levels will have a direct effect. Yet in Supplemental Table 3 we see there is more FI than FH in VH (608ng/ml vs 466ng/ml respectively). Therefore, adding more FI without more co-factors would have a very limited effect. Surely this demonstrates that the study was delivering the wrong payload, i.e. FI, which hit a natural ceiling of endogenous co-factors within the eye?

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

This study analyzed biomarker data from 28 subjects with geographic atrophy (GA) in a Phase I/II clinical trial of PPY988, a subretinal AAV2 complement factor I (CFI) gene therapy, to evaluate pharmacokinetics and pharmacodynamics. Post-treatment, a 2-fold increase in the vitreous humor (VH) FI was observed, correlating with a reduction in FB breakdown product Ba but minimal changes in other complement factors. The aqueous humor (AH) was found to be an unreliable proxy for VH in assessing complement activation. In vitro assays showed that the increase in FI had a minor effect on the complement amplification loop compared to the more potent C3 inhibitor pegcetacoplan. These findings suggest that PPY988 may not provide enough FI protein to effectively modulate complement activation and slow GA progression, highlighting the need for a thorough biomarker review to determine optimal dosing in future studies.

Strengths:

This manuscript provides critical data on the efficacy of gene therapy for the eye, specifically introducing complement FI expression. It presents the results from a halted clinical trial, making sharing this data essential for understanding the outcomes of this gene therapy approach. The findings offer valuable insights and lessons for future gene therapy attempts in similar contexts.

Weaknesses:

No particular weaknesses. The study was carefully performed and limitations are discussed.

I have just some concerns about the methodology used. The authors use the MILLIPLEX assays, which allow for multiplexed detection of complement proteins and they mention extensive validation. How are the measurements with this assay correlating with gold standard methods? Is the specificity and the expected normal ranges preserved with this assay? This also stands for the Olink assay. Some of the proteins are measured by both assay and/or by standard ELISA. How do these measurements correlate?

The authors thank the reviewer for the positive response. Regarding the ELISA assays used to measure the array of complement proteins described, these were extensively validated for the following parameters: specificity, intra-assay and inter-assay precision, accuracy, stability, reference range, and parallelism. All assays were validated in plasma, vitreous and aqueous humour. Due to the limited volume and availability of ocular fluids from individuals in the study, validation in vitreous and aqueous matrices was performed using a pool of several samples from post-mortem donors. At the time this study was initiated, the Millipore Luminex complement panels and the Quidel C3a and Ba EIA were the most sensitive assays and the only commercially available options capable of measuring the proteins of interest in the context of limited vitreous and aqueous humor sample. The concentrations measured were observed at similar ranges as those published in the literature using assays in distinct patient populations e.g. in (Mandava et al, Invest Ophthalmol Vis Sci, 2020).

Measurements from vitreous and aqueous from subject samples were deemed reportable if they were within the quantifiable ranges defined for these sample types during the validation (coefficient of variation of 20%, or 30% when results were below the lower limit of quantification but above limit of detection). Notably, given the limited amount of biomarker data due to small sample size, we share results from outlier biomarker measurements, to illustrate the heterogeneity in sample quality. We further publish plasma sample biomarker results in supplemental table 5 wherein complement protein concentrations can be observed and compared to normal ranges in the literature.

Adding confidence to the robustness of our assays was the observation that some of the complement proteins quantified by standard assay (e.g. plate and bead-based ELISAs) were also measured by the OLINK assay, and there was a general trend observed for positive correlation between results from both assays for FI levels post-treatment. However, we did not provide detailed correlative statistical analyses for further complement proteins as OLINK findings were deemed highly exploratory and hypothesis generating, and because the OLINK assay produced normalised results which are challenging to directly compare to ELISA results that were absolute.

Reviewer #2 (Public Review):

Summary:

The results presented demonstrate that AAV2-CFI gene therapy delivers long-term and marginally higher FI protein in vitreous humor that results in a concomitant reduction in the FB activation product Ba. However, the lack of clinical efficacy in the phase I/II study, possibly due to lower in vitro potency when compared to currently approved pegcetacoplan, raises important considerations for the utility of this therapeutic approach. Despite the early termination of the PPY988 clinical development program, the study achieved significant milestones, including the implementation of subretinal gene therapy delivery in older adults, complement biomarker comparison between serial vitreous humor and aqueous humor samples and vitreous humor proteomic assessment via Olink.

Strengths:

Long-term augmentation of FI protein in vitreous humor over 96 weeks and reduction of FB breakdown product Ba in vitreous humor suggests modulation of the complement system. Developed a novel in vitro assay suggesting FI's ability to reduce C3 convertase activity is weaker than pegcetacoplan and FH and may suggest a higher dose of FI will be required for clinical efficacy. Warn of the poor correlation between vitreous humor and aqueous humor biomarkers and suggest aqueous humor may not be a reliable proxy for vitreous humor with regard to complement activation/inhibition studies.

Weaknesses:

The vitrectomy required for the subretinal route of administration causes a long-term loss of total protein and may influence the interpretation of complement biomarker results even with normalization. The modified in vitro assay of complement activation suggests a several hundred-fold increase in FI protein is required to significantly affect C3a levels. Interestingly, the in vitro assay demonstrates 100% inhibition of C3a with pegcetacoplan and FH therapeutics, but only a 50% reduction with FI even at the highest concentrations tested. This observation suggests FI may not be rate-limiting for negative complement regulation under the in vitro conditions tested and potentially in the eye. It is unclear if pharmacokinetic and pharmacodynamic properties in aqueous humor and vitreous humor compartments are reliable predictors of FI level/activity after subretinal delivery AAV2-CFI gene therapy.

The authors thank the reviewer for the positive response and we agree that a limitation of the biomarker strategy for ocular gene therapy delivered to the retinal tissues is inferring PK/PD from vitreous and aqueous samples, which are the fluid sample compartments accessible from subjects available to measure molecular treatment response. We agree that these compartments may not accurately represent sub-retinal and tissue level complement turnover. In the discussion, line 508, we state: ‘Overall, the data suggests that fully functional FI is being secreted into the VH, but the regulatory effects on the level of Ba may be representative of convertase formation in the VH and not the macula retina/RPE nor the choroid. To validate this hypothesis, one approach would be to conduct vitreal sampling using an effective drug targeting C3 for GA in a larger cohort’.

However, the observation of elevation of FI in VH (and AH) post treatment, and changes in levels of downstream complement proteins that align with prior knowledge of control of alternative pathway activation, is compelling evidence that these measurements reflect modest but direct consequences of an FI-gene therapy that was delivered to the subretinal space. We add to the discussion, line 479: ‘the findings of elevated FI in the VH after sub-retinally delivered CFI gene therapy and changes in complement pathway proteins post-treatment build confidence that VH matrix is at least partially reflecting the complement system at the retinal layers and treatment site, and is a valid biomarker for PK/PD insights in response to treatment.’

Furthermore, the observation of moderately raised FI levels in modelled VH post treatment being insufficient to control CS activation in vitro accords with the lack of clinical response observed at phase II. We note that measuring FI and complement biomarkers in retinal tissues from treated eyes at post-mortem would be one way to explore the PK/PD effects from AAV2-FI gene therapy.

Reviewer #3 (Public Review):

Summary:

The manuscript by Hallam et al describes the analysis of various biomarkers in patients undergoing complement factor I supplementation treatment (PPY988 gene therapy) as part of the FOCUS Phase I/II clinical trial. The authors used validated methods (multiplexed assays and OLINK proteomics) for measuring multiple soluble complement proteins in the aqueous humour (AH) and vitreous humour (VH) of 28 patients over a series of time points, up to and including 96 weeks. Based on biomarker comparisons, the levels of FI synthesised by PPY988 were believed to be insufficient to achieve the desired level of complement inhibition. Subsequent comparative experiments showed that PPY988-delivered FI was much less efficacious than Pegceptacoplan (FDA-approved complement inhibitor under the name SYFORVE) when tested in an artificial VH matrix.

Strengths:

The manuscript is well written with data clearly presented and appropriate statistics used for the analysis itself. It's great to see data from real clinical samples that can help support future studies and therapeutic design. The identification that complement biomarker levels present in the AH do not represent the levels found in the VH is an important finding for the field, given the number of complement-targeting therapies in development and the desperate need for good biomarkers for target engagement. This study also provides a wealth of baseline complement protein measurements in both human AH and VH (and companion measurements in plasma) that will prove useful for future studies.

Weaknesses:

Perhaps the conclusions drawn regarding the lack of observed efficacy are not fully justified. The authors focus on the hypothesis that not enough FI was synthesised in these patients receiving the PPY988 gene therapy, suggesting a delivery/transduction/expression issue. But beyond rare CFI genetic variants, most genetic associations with AMD imply that it is a FI-cofactor disease. A hypothesis supported by the authors' own experiments when they supplement their artificial VH matrix with FH and achieve a significantly greater breakdown of C3b than achieved with PPY988 treatment alone. Justification around doubling FI levels driving complement turnover refers to studies conducted in blood, which has an entirely different complement protein profile than VH. In Supplemental Table 5 we see there is approx. 10-fold more FH than FI (533ug/ml vs 50ug/ml respectively) so increasing FI levels will have a direct effect. Yet in Supplemental Table 3 we see there is more FI than FH in VH (608ng/ml vs 466ng/ml respectively). Therefore, adding more FI without more co-factors would have a very limited effect. Surely this demonstrates that the study was delivering the wrong payload, i.e. FI, which hit a natural ceiling of endogenous co-factors within the eye?

See response to reviewer 3’s review after reviewer 3 recommendations section below.

Recommendations for the authors:

Reviewer #2 (Recommendations For The Authors):

The authors present strong evidence using validated complement biomarker assays and comprehensive proteomic profiling that support their findings. The presentation of complement biomarker data in vitreous humor and aqueous humor after FI augmentation is presented in a clear and concise format. The direct comparison of complement biomarkers in vitreous humor and aqueous humor from the same patients and demonstrating similarities and differences is important for the nascent complement gene therapy field. Developing a novel in vitro complement model and comparing pegcetacoplan, FH, and FI inhibitors provides the field with a valuable assay to benchmark other complement therapeutics. As currently designed, the in vitro assay supports why FI augmentation did not contribute to clinical success. It also suggests that non-physiological concentrations of FI protein (over 100 µg/mL) maximally inhibit C3a signal by ~50%, whereas both pegcetacoplan and FH reduce the signal by 100%. Does this suggest that CFI is not an appropriate therapeutic target to control complement overactivation in the eye?

We agree with the reviewer that the new data from the novel in vitro assay coupled with the clinical findings from the phase II gene therapy trial does now suggest FI is less attractive as a therapeutic target for controlling complement activation in the retinal tissues of subjects with Geographic Atrophy.

Reviewer #3 (Recommendations For The Authors):

I think the authors have done a great job collecting and analysing these clinical samples and elucidating the baseline complement protein profile in both the AH and VH. I only have minimal suggested changes.

Perhaps a more direct discussion around the limitations of adding more FI into environments where there is no excess of FI-cofactors present? And a discussion around the limitations of VH (and VA for that matter) biomarker sampling for a disease that primarily affects the neurosensory retina and outer blood/retinal barrier: perhaps the landscape of complement proteins is different yet again (although, admittedly, impossible to sample in a patient)? Finally, would it not have been better to perform complement activation experiments using the VH of treated patients directly rather than creating an artificial VH matrix (there may, or may not, be a couple of things in human VH that directly affect complement turnover...)?

We thank the reviewer for the supportive comments. This study is the first to describe FI and FH levels and respective ratios in vitreous humour (plus aqueous and plasma) from GA subjects, before and after sub-retinal gene therapy. It is compelling to observe that in the VH the levels of FI are greater than FH, the primary fluid phase co-factor for FI enzymatic activity. This new information does indeed argue against further FI supplementation (using gene therapy) being of added benefit to controlling the complement system in the broader population in individuals with Geographic Atrophy. We note that at the start of the clinical development of GT005/PPY988 AAV2-FI gene therapy, there was limited information on FI and FH levels in AMD in ocular fluids to inform the pharmacodynamics of complement activation. Now, by running the FOCUS phase I clinical trial and measuring the complement biomarker data using validated assays we have added to our understanding on the levels and ratio of FI to FH and other complement proteins in a larger number of GA subjects’ ocular samples. We report the levels of complement proteins measured in ocular and systemic samples, to show the ranges and also the differences in ratios between the different matrices.

Regarding the statement that FI supplementation could likely be ineffective due to limited FH cofactor; FH is not the only co-factor that FI may partner with at cell surfaces to become enzymatically active (others include MCP (CD46) and CR1 (CD35), although the latter is known to be of limited expression in the eye), as such, it is certainly true that other proteins may be present in the tissue altering the kinetics of FI’s activity after sub-retinal gene-therapy. In addition, the ratio between FI and FH detected in the VH may not be the same as in retinal tissue. As such, we agree that drawing insights from biomarkers in the VH may not fully reflect the disease processes and treatment response at the retinal cell layers, but it is the closest fluid sample available to sample tissue released soluble proteins. We acknowledge that VH biomarkers will not fully capture retinal disease processes and treatment responses, but due to their proximity, will reflect retina-released soluble proteins. The findings of elevated FI in the VH after sub-retinally delivered CFI gene therapy and changes in complement pathway proteins post-treatment build confidence that VH matrix is at least partially reflecting the complement system at the retinal layers and treatment site, and is a valid biomarker for PK/PD insights in response to treatment. We agree modelling different inhibitor effects on complement activation directly using subject’s vitreous would be informative, but this was not possible due to the limitations of very small sample volume.

We add several sentences to the discussion regarding the points above. Line 473: ‘Notably, that FI does not reduce C3a breakdown to baseline even at supermolecular concentrations suggests cofactor limitation that might be more pronounced in VH given FH is not in excess of FI as is the case in blood 27. Moreover, there are additional cell-bound cofactors for FI that may be present in retinal tissue that are not present in the VH and could further alter the kinetics of the assay, such as MCP (CD46) albeit with disease related changes observed 37. However, the findings of elevated FI in the VH after sub-retinally delivered CFI gene therapy and changes in complement pathway proteins post-treatment build confidence that VH matrix is at least partially reflecting the complement system at the retinal layers and treatment site, and is a valid biomarker for PK/PD insights in response to treatment.’

Minor comments:

Line 237: Missing parenthesis at the end of the sentence

Manuscript updated.

Line 435: Missing secondary parenthesis after .....Figure 3A)......

Manuscript updated.

Line 536: I don't think suggesting the addition of FHR proteins into the neurosensory retina/VH is such a good idea

The reference to FHRs has been clarified in the manuscript, line 558. The authors note that FHR dimerization domains have been engineered to dimerize Factor H constructs increasing half-life and potency for drugs currently in development.

Version 2

Abstract

Abstract

Objective

Complement biomarker analysis in ocular fluid samples from subjects with geographic atrophy (GA) in a Phase I/II clinical trial of subretinal AAV2 complement factor I (CFI; FI) gene therapy, PPY988 (formerly GT005), to understand target pharmacokinetics/pharmacodynamics. Clinical findings were subsequently utilized to investigate the therapeutic dose in an in vitro complement activation assay.

Design, setting and participants

Biomarker data were evaluated from 28 subjects in FOCUS, a Phase I/II clinical trial evaluating the safety and efficacy of three ascending doses of PPY988.

Main outcomes and measures

Vitreous humor (VH), and aqueous humor (AH) from subjects before surgery and at serial timepoints (week 5 or 12, 36, 96) were evaluated for changes in levels of intact complement factors I, B and H (FI, FB, FH) components C3, C4, and C1q and breakdown products (Ba, C3a, C3b/iC3b, C4b) using validated assays and OLINK® proteomics.

A modified in vitro assay of complement activation modelling VH complement concentrations was used to compare PPY988 potency to the approved intravitreal C3 inhibitor pegcetacoplan (Apellis) and complement Factor H (FH).

Results

An average 2-fold increase in VH FI was observed post-treatment at week 36 and week 96. This correlated with a marked post-treatment reduction in VH concentration of the FB breakdown product Ba and Ba:FB ratio, but minimal changes in C3a and C3b/iC3b levels. Variable concordance in complement biomarker levels in VH versus AH suggest AH is not a reliable proxy for VH for complement activation. During the experimental comparison of doses, a 2-fold increase of FI achieved in the vitreous had only a minor effect on the complement amplification loop in vitro, indicating limited impact [IC50: 1229nM]. Pegcetacoplan completely blocks C3a generation at concentrations much lower than the estimated trough level for monthly intravitreal injections [IC50: 2nM]. Supplementation with FH in the assay revealed similar potency to pegcetacoplan [IC50: 6nM].

Conclusions and relevance

PPY988 subretinal gene therapy may not have provided sufficient FI protein to meaningfully modulate complement activation to slow GA growth. Reviewing VH biomarkers is important for understanding target expression, pathway engagement, and determining optimal dose, thereby informing future clinical development.

Introduction

Age-related macular degeneration (AMD) is a leading cause of blindness, particularly in developed countries. In the United States (US) alone, there are approximately 18 million individuals aged 40 and above with early-stage AMD, and 1.5 million with late-stage AMD, which includes geographic atrophy (GA) and/or neovascular AMD 1. GA is a condition characterized by the progressive loss of photoreceptors, retinal pigment epithelium (RPE), and underlying choriocapillaris, resulting in a central scotoma 2. Reduced vision can significantly impact activities of daily living and quality of life 3,4. There have been two recent Food and Drug Administration (FDA)-approved therapies that inhibit complement activity, with a modest slowing of GA growth 5,6. However, they require life-long monthly or bimonthly intravitreal (IVT) injections and have so far exhibited limited functional efficacy that present challenges for long term compliance 7. Therefore, a one-time gene therapy that provides long-lasting efficacy equivalent to or better than repeated IVT therapies is a promising avenue for exploration.

One such gene therapy, called PPY998 (formerly GT005), is a subretinal gene therapy that supplements complement factor I (CFI; FI) protein in the retinal layers where atrophy occurs 8. A Phase II study (HORIZON, NCT04566445) was initiated, however a decision to stop the PPY988 program was taken by Novartis based on a recommendation from the independent Data Monitoring Committee following an overall benefit-risk assessment of available data from the program studies. [Press release: Novartis AG; September 11th 2023, : https://www.novartis.com/news/gt005-ppy988-development-program-geographic-atrophy]. This raises important questions about whether CFI was sufficiently expressed after subretinal gene therapy, whether there was evidence of modulation of intraocular complement activity, or if this modulation was sufficient to potentially alter the disease course. Such information is crucial to the retina community, as there are many companies developing complement modulating therapies for AMD 9. Sharing the learnings from the PPY998 program would improve the risk-benefit analysis for these programs and help understand if new approaches differentiate from PPY998, justifying further clinical study of complement targets for GA.

The complement system, specifically the alternative pathway (AP), plays an important role in the development of AMD, and increasing local FI concentration could be a promising approach to regulate ocular complement activity. The AP is a part of the complement system that constantly surveys for foreign cells and debris and requires tight regulation to prevent inflammation and self-cell damage 10. Regulatory proteins like factor H (FH), decay-accelerating factor (DAF; CD55), membrane cofactor protein (MCP; CD46), complement receptor 1 (CR1; CD35), and FI help control the AP, with FI being the only regulatory enzyme 11. FI, expressed from the CFI gene, cleaves C3b by forming a trimolecular complex with cofactors FH for C3b, and MCP and CR1 for C3b and C4b 12–18. Studies have shown that carriers of rare variants (RV) in CFI have increased risk of developing advanced AMD and more rapid disease progression 19–23. CFI RVs can result in low levels of serum FI (Type I variants) or normal levels with functional impairment (Type II variants) 24,25, many examples of which were observed in Gyroscope’s SCOPE natural history study 26. Research has also suggested that doubling serum FI levels may help control complement overactivation driven by other AMD risk genetic factors mapping to complement genes 27. Based on these findings, clinical trials were conducted to explore the potential of sub-retinal delivery of AAV2-CFI as a gene therapy for treating dry AMD 8.

The FOCUS study is a Phase I/II open-label study evaluating the safety, and dose response of three doses of PPY988 AAV2-CFI subretinal gene therapy in GA. To assess transgene pharmacokinetics and pharmacodynamics, vitreous humor (VH) and aqueous humor (AH) samples were collected at baseline (BL) and four time points up to 96 weeks after dosing and were evaluated for exploratory complement biomarkers. This report shares findings regarding change from baseline in FI and complement protein levels and activation, over time after vitrectomy and PPY988 administration. Levels of FI achieved in VH from FOCUS subjects were then used in an in vitro simulated human vitreous complement activation assay, to gain insights on the therapeutic potencies of FI, C3 inhibitor pegcetacoplan (Syfovre, Apellis), and FH. We also offer critical learnings for other AMD complement gene therapy programs and a guide for future analyses of complement biomarker data in ocular samples.

Methods

Ethics

The FOCUS study was conducted in accordance with the tenets of the Declaration of Helsinki and all other local regulations, in accordance with the International Conference on Harmonization E6 Guidelines for Good Clinical Practice and with applicable local, state, and federal laws. Institutional review board or ethics committee approval was obtained, and the study was registered at ClinicalTrials.gov (reference: NCT03846193) and EudraCT (reference: 2017-003712-39). All subjects provided written informed consent to participate, and an independent data monitoring committee provided ongoing oversight.

Study design

FOCUS is an open-label first in human Phase I/II multicentre clinical trial to evaluate the safety, and dose response of PPY988 (AAV2-CFI) 8, administered as a single subretinal injection in subjects with GA secondary to AMD, followed for up to 240 weeks (5 years) post-dosing. Subjects were administered one of three doses to the study eye under local anaesthesia by a qualified vitreoretinal surgeon: 2E10 viral genome (vg; low), 5eE10 vg (medium) and 2E11 vg (high). At screening, a saliva sample was taken from subjects and extracted DNA was analysed for CFI RV using targeted next generation sequencing 21. Genotyping results were not required at screening if already available through participation in a previous Gyroscope sponsored study. The term ‘rare’ was defined as variants that changed the CFI coding sequence, had ≤1% minor allele frequency in GnomAD non-Finnish Europeans (V2.1.1, accessed http://gnomad.broadinstitute.org), and were previously associated with FI haploinsufficiency 11. A detailed description of study design, route of administration, dose, number of subjects, biomarker matrices, sampling timing and methodology are provided in Supplementary Methods and Supplementary Table 1.

Complement protein and total protein quantification

AH and VH samples of subjects were taken on the treated eye immediately prior to subretinal surgery at baseline (BL) and at 5 or 12 weeks, 36, and 96 weeks post-treatment, as described in detail in Supplementary Table 1. Samples were rapidly frozen at −80 °C and transported frozen for analysis. Prior to analysis, samples were thawed and subjected to centrifugation at 17,000g for 10 minutes at 4°C.Samples were kept on ice during sample preparation.

Complement factors C2, C4b, C5, C5a, factor D, mannose binding lectin (MBL), FI, C1q, C3b/iC3b, C3, FH, FB and C4 were quantified in AH and VH using the MILLIPLEX Human Complement Panel 1 and 2 (cat. no. HCMP1MAG-19K and HCMP2MAG-19K, EMD Millipore Corporation, Billerica, MA) following extensive quality control and validation of calibration curves. Plates were read using the Luminex MagPix and xPONENT software (Luminex Corporation, Austin, TX). C5a was omitted from further analysis as measurements fell under the lower limit for quantification.

Complement factors C3a and Ba were quantified in AH and VH using MicroVue EIA kits (cat. no. A032 and A034, QuidelOrtho Corporation, San Diego, CA) following manufacturer instructions. Plates were read using a Tecan Sunrise absorbance microplate reader using the Magellan Tracker data analysis software (Tecan Trading AG, Switzerland).

Each assay was validated and considered fit for purpose. Sample processing and analyses were conducted in accordance with Good Clinical Laboratory Practice, the analytes or measurements that did not meet acceptance criteria (coefficient of variation of 20%, or 30% when results are below the lower limit of quantification but above limit of detection) were excluded from the final analysis. The number of successfully sampled subjects for each biomarker at each time-point is reported in Supplementary Table 2.

Total protein was quantified in AH and VH using the Quant-iT protein assay kit (cat. no. Q33210, Invitrogen, Thermo Fisher Scientific, Waltham, MA) as per the manufacturer’s instructions. Plates were read using a Varioskan LUX fluorescent microplate reader using the SkanIt data acquisition software (Thermo Fisher Scientific, Waltham, MA).

In vitro modelling the ocular complement system and potency assay

The assay was performed using reagents from the Wieslab AP assay kit (cat. no. COMPLAP330, Svar Life Science AB, Sweden) and purified human serum complement proteins (Complement Technology Inc., Texas USA). All preparation steps were performed on ice. Serial dilutions of the test analytes FI, FH and pegcetacoplan were separately prepared and added to a mix of the complement components C3 (2543.3ng/mL), FB (853.6ng/mL), FH (296.2ngmL), FI (603.8ng/mL) and Properdin (18.1ng/mL) simulating the final median concentrations measured in the VH baseline of FOCUS subjects (concentrations of C3, FB, FH, FI are from CFI RV negative population, Properdin concentration is from a limited number of FOCUS subjects from both populations). The simulated vitreous complement mix was transferred on the lipopolysaccharide (LPS) coated Wieslab plate on ice. FD was separately added to the mix, the plate was immediately put at 37°C and incubated for 1 hour. At the end of the incubation, the solution was transferred into a dilution plate and kept on ice, EDTA was added to each well to obtain a final 20mM concentration. Ba and C3a generated in the assay were measured by using the MicroVUE EIA kits (cat. no. A032 and A034, QuidelOrtho Corporation, San Diego, CA) following the manufacturer instructions. The ELISA plates were scanned using the Varioskan plate reader and analysed using the SkanIT software (Thermo Fisher Scientific, Waltham, MA). The in vitro FI potency comparability assay is described in Supplementary Methods. The trough level of pegcetacoplan in the VH after 27 days was calculated by halving the treatment dose of 15mg/eye 6 times (pegcetacoplan half-life 4.5 days), the number obtained was converted to µg/mL using an estimated volume of VH of 4 mL 28.

Statistics

All subject sample biomarker analyses were performed in R version 4.1.0. The analysis set consisted of all subjects who received subretinal delivery of PPY998 from cohorts 1 to 4 and had at least one non-missing biomarker assessment performed at BL and either week 36 or week 96. Given sparsity of data at week 5 and week 12, the data were combined, and the average value was used in instances where subjects had non-missing data from both visits. Concentrations values above the upper limit of quantification (ULOQ) was set to ULOQ and values below the lower limit of detection (LOD) were set to LOD. Please see Supplementary Tables 3-4 for limits.

Significance testing for the change in protein level from BL to week 36 or week 96 were performed on the differences of log-transformed data with the non-parametric Wilcoxon rank-sum test. Means and standard error bars are shown in the respective figures. Correlations between VH and AH sampling, as well as between FI and other proteins, were summarized with the Spearman’s Rank correlation coefficient. As analyses were exploratory in nature and hypothesis-generating, no adjustments for multiple testing were performed.

For the in vitro VH modelling assay, to calculate IC50s, 4-parameter logistic curves were generated using GraphPad Prism version 10.0.0 for Windows, GraphPad Software, Boston, Massachusetts USA, www.graphpad.com.

Results

Biomarker detection in FOCUS subjects that received PPY988

In FOCUS Part 1 and 2 (cohorts 1 to 4), 31 subjects received one of three doses of PPY988 via the standard subretinal delivery approach (Supplementary Table 1). A total of 97% of the ocular samples expected under the protocol were obtained prior to early termination of the study which triggered cessation of further ocular fluid sampling. A total of 88 VH, 63 AH and 45 plasma samples from 30 subjects were obtained at various timepoints including BL, and 5 or 12, 36 and 96 weeks post-treatment (Supplementary Methods and Supplementary Tables 1 and 2).

In FOCUS, PPY988 was well tolerated with no serious ocular AEs, some dose-related RPE changes in the bleb area were evident and there was no clinically significant vector-related inflammation. A full account of the FOCUS study safety and clinical data, as well as full Phase II PPY988 safety, efficacy and biomarker data will be reported separately following study completion.

Vitrectomy for subretinal AAV-CFI gene therapy procedure results in sustained loss of protein from the VH

To test impact of total vitrectomy performed during transvitreal subretinal delivery of PPY988 (Figure 1A and B) influencing global proteomic matrix changes and biomarker analyses/interpretation, total protein (TP) measurements were obtained at BL and each sampling timepoint (according to sample availability). In VH, a ∼30% reduction in TP from baseline was observed at either week 5 or 12 and week 36 (NS, P=0.162), with a significant >50% reduction identified at week 96 (P=0.033), while there were minimal changes in AH TP at week 5/12, 36 or 96 (Figure 1C). Because VH TP was reduced post-operatively, we performed TP normalisation to compare the results with non-normalised data interpretation for VH but not AH.

Surgical delivery of PPY988 AAV2-CFI. (A) PPY988 vector design. FI protein is secreted by cells targeted by AAV2 serotype and containing the cDNA sequence for wild-type CFI (NM_000204.4) driven by a CAG promoter 8. (B) Subretinal delivery of PPY988. (C) Vitrectomy during SR delivery of PPY988 results in loss of total protein (TP) in the vitreous humor. TP concentration (y-axis, log2 µg/µL) was measured at baseline and 5 or 12, 36, and 96 weeks post-surgery (y-axis). Mean +/− SE is shown by blue bars and lines whilst all subjects are shown by black lines for both vitreous (left) and aqueous (right) humor. A significant reduction in TP was observed at week 96 in vitreous humor (*P=0.033). Significance is defined as *P<0.05. BL = Baseline; WK= week.

AAV2-CFI delivery results in a sustained increase in FI protein level in the VH and AH

Transgene expression was measured in VH and AH samples at weeks 5 or 12, 36, and 96 after treatment with PPY988. On average, we identified a 2-fold increase in VH FI at week 36 compared to baseline, with non-TP normalized baseline concentrations of 500 ng/mL compared to 1000 ng/mL at week 36 (non-normalised P<0.001; TP normalised increase ∼2-fold (P=0.002)) (Figure 2A, 2B). The increase in FI concentration was sustained, with a ∼2.5-fold increase in FI observed at week 96. (mean ∼1250 ng/mL) (non-normalised P<0.001; TP normalised increase ∼3-fold (P=0.002)). CFI RV (n=5) and non-CFI RV (‘broad’) (n=19) genotype groups demonstrate comparable magnitudes of absolute FI elevation post-treatment, with the overall CFI RV cohort FI starting concentration ∼50% reduced (NS). Elevated FI expression in AH samples at week 36 (P<0.001) were sustained out to week 96 (P<0.001) (Figure 2C, 2D). Lastly, there was a variable response to the three PPY988 doses in both VH and AH with no significant dose effect observed (Figure 2C, 2D).

Temporal expression of factor I (FI) in vitreous humor (VH) and aqueous humor (AH) after PPY988 delivery. (A,B) FI levels (y-axis, ng/mL) in VH were significantly increased at week 36 and 96 post-surgery (x-axis) both when non-normalised (A) (P=0.001 for both timepoints) and total protein -normalised (B) (P=0.002 at both timepoints). Mean +/− SE is shown by bars for all (black lines) broad (blue lines) and CFI RV (green lines) groups. (C,D) FI level (y-axis, ng/mL) for each subject in the cohort is tracked by week (x-axis) and separated by dose (2E10 vg: orange lines; 5E10 vg: green lines; 2E11 vg: blue lines) in VH (C) and AH (with significant increase in AH FI at week 36 and 96 (both P<0.001)) (D), where mean +/− SE shown by black bars. Significance is defined as *P<0.05; **P<0.01, ***P<0.001. N numbers are shown above X-axis. BL = Baseline; WK= week.

Increased FI expression results in a reduction in Ba and Ba:FB in the VH

To test for FI target engagement at the level of formation of the alternative pathway C3 convertase, we measured Ba and FB concentrations at each sampling week after subretinal delivery of PPY988. A reduction in Ba in the VH is indicative of a reduction in C3 convertase formation in the ocular environment. We observed that Ba levels were reduced by ∼2-fold compared to baseline at week 36 and 2.5-fold at week 96 in non-normalised data (week 36: P=0.004; week 96: P=0.003) (Figure 3A) and TP normalised data (week 36: P=0.002; week 96: P=0.038) (Figure 3B). To add confidence to this read-out by accounting for fluctuations in endogenous FB biosynthesis, ratios of intact complement proteins and their breakdown products were assessed as in previous studies 29–31. Ba:FB ratio changes post-treatment reflected the raw data. When reviewing biomarker levels from both non-normalized and TP normalized datasets, a significant reduction in Ba:FB was observed at week 36 (both P<0.001) and at week 96 (non-TP normalized (P=0.001); TP normalized (P=0.033)) (Figure 3C, 3D).

Modulation of Ba and Ba:FB by PPY988 delivery. (A,B) Ba concentrations (Y-axis, ng/mL) in vitreous humor (VH) samples by week post-surgery (X-axis). Mean +/− SE is shown by bars for all (black lines) broad (blue lines) and CFI RV (green lines) groups. Significant reductions in Ba were observed at week 36 for both non-normalised (P=0.004) (A) and total protein (TP) normalised (P=002) (B) data and week 96 for non-normalised data (P=0.003) and TP normalised (P=038). (C,D) Ratios of Ba:FB concentration (y-axis, ng/mL) were calculated for each subject at each sampling post-surgery week (x-axis). Mean +/− SE is shown by bars for all (black lines) broad (blue lines) and CFI RV (green lines) groups. Significant reductions were identified at week 36 for both non-normalised (P<0.001) (C) and TP normalised (P<0.001) (D) Ba:FB ratios, with a reduction at week 96 for non-normalized (P=0.001) and TP normalized (P=0.033) ratios. Significance is defined as *P<0.05; **P<0.01, ***P<0.001. N numbers are shown above X-axis. BL = Baseline; WK= week.

Increased FI expression results in modest alterations in VH and AH complement biomarker profiles

To investigate the effect of PPY988 delivery on complement AP and classical pathway (CP) activity, a panel of biomarkers including FI, C3, FB, FH, C1q, Ba, C3a, C3b/iC3b, C4 and C4b were measured in the VH and AH (in addition to plasma) at BL, week 5 or 12, week 36 and week 96. The complete set of concentrations at BL are reported in the Supplementary Table 3 (VH), Supplementary Table 4 (AH), and Supplementary Table 5 (plasma). In addition, Ba:FB, C3a:C3 and C3b/iC3b:C3 ratios were calculated at each time-point. Baseline values for each biomarker with descriptive statistics are reported in Supplementary Table 3 (VH) and Supplementary Table 4 (AH), with baseline and post-treatment pairwise correlations between FI and biomarkers presented in Supplementary Figure 1 (VH) and Supplementary Figure 2 (AH). Fold changes in complement biomarker concentration from baseline are presented for each biomarker or ratio at each post-surgery sampling week in VH non-normalised data (Figure 4A), VH TP normalised data (Figure 4B) and AH non-normalised data (Figure 4C). In the non-normalised VH dataset, we observed reductions in Ba and Ba:FB, with a reduction in C1q possibly indicative of reduced ocular immune cell (e.g. microglial) activation 32 (Figure 4A). We detected no modulation of VH C3a and C4b. Although levels of C3b/iC3b per unit of C3 fall substantially at week 96 in both non-normalized (P=0.002) and TP-normalized (P=0.012) analyses (Supplementary Figure 3), pairwise correlation analyses reveal that no biomarkers or ratios were negatively correlated with increasing FI level at week 36 or 96 post-treatment (Supplementary Figures 1 and 2). Minimal reduction of CP activity was indicated by C4b:C4 ratio analysis, however, this was not observed when normalizing for TP (Supplementary Figure 4). VH FB and FH concentrations do not fluctuate substantially from BL values after treatment with PPY988. Overall, TP normalisation did not substantially alter the interpretation of VH biomarkers with broadly similar trends observed (Figure 4B).

Modulation of complement AP and CP biomarkers by PPY988 delivery. Fold-changes (y-axis) in complement biomarker [FI, C3, FH, FB, C1q, Ba, C3a, C3b/iC3b, C4b, Ba:FB, C3a:C3, C3b/iC3b:C3] concentrations from baseline at each post-surgery week (VH: week 5/12, 36 and 96; AH: week 36 and 96; X-axis) are shown in VH with no normalisation (A), VH with total protein normalisation (B), and AH with no normalisation. Mean is shown by thick black lines, +/− SE is shown by bars. Individual subject changes are shown by grey lines. Yellow boxes highlight disparate trends in VH vs AH biomarkers. BL = Baseline; WK= week.

AH sampling is not reflective of VH complement activation profile

Complement biomarker profiling in AH demonstrated increasing concentrations of a majority of biomarkers (Figure 4C), including for Ba, C1q, and C3a (at week 96) which was discordant with VH sampling data. Whether these components (C1q) and breakdown products (Ba, C3a) are generated locally, systemically, or diffusely from the retina or VH is unclear. A similar trend to VH was observed when considering the increase in AH C3 and associated fall in C3a:C3 and C3b/iC3b:C3 ratios at week 96 in particular. However, in the case of AH, the increase in C3 was also observed at week 36, suggesting compartmentalization of response of the AH vs the VH. In line with these data, pairwise correlations between VH and AH levels of Ba, C3a and C3b/iC3b post-treatment are modest and variable (Supplementary Figure 5), with only FI consistently well correlated between the two ocular compartments.

Proteomic analysis on VH using Olink provides supportive evidence for target expression and modest modulation of the complement system

Olink proteomic analysis of the VH samples identified a 2-fold median increase in week 36 FI levels relative to baseline (Supplementary Methods, Supplementary Figure 6a). Further review of differentially expressed proteins revealed trends for elevation of additional complement components including factor B, properdin and C2 (Supplementary Figure 6b, Supplementary Table 6).

VH modelling of complement activation reveals inferior potency of FI vs pegcetacoplan

By modelling the components of the VH based on median measurements of C3, FB, FH, FI (broad cohort measurements) and properdin (all measurements) at baseline in the FOCUS study, we compared the potency of FI, FH and the C3 inhibitor pegcetacoplan for their effectiveness in controlling the generation of C3a on an LPS-coated plate in a simulated ocular environment context (Figure 5). In this assay, C3 is directly activated by O-antigens on LPS thus triggering activation to drive the AP amplification loop. A wide range of FI, FH and pegcetacoplan concentrations were titrated through an artificial VH matrix so that comparisons could be made to clinically relevant levels of both FI and pegcetacoplan. Serum purified FI was revealed to be of similar potency to FI expressed from Good Manufacturing Practice (GMP) manufactured PPY988 (Supplementary Figure 7). The trough level of pegcetacoplan at 27 days was estimated using injection dose and half-life information available on the SyfovreTM prescribing information datasheet. No diffusion data in human was available so this was not counted in the trough calculation. The concentration obtained was similar to that reported in the pegcetacoplan rabbit pharmacokinetics (PK) study and PK/PD prediction study published by Apellis 33,34. Pegcetacoplan vitreal trough level was compared to the 1st and 3rd quartile increase (at week 36 vs BL) in FI associated with PPY988 delivery. Pegcetacoplan exhibited greater potency than FI, with complete inhibition of the AP amplification loop and C3a generation at predicted trough level (∼59 μg/mL), whilst FI was observed to have minimal effect at levels of expression associated with PPY988 after 36 weeks. Overall curves generated by logistic regression revealed an IC50 of 1229 nM for FI, whilst FH [IC50: 6 nM] was similar in potency to pegcetacoplan [IC50: 2 nM]. The estimated vitreous trough level of pegcetacoplan was approximately 100-fold greater than that required for complete inhibition of the AP in this assay. On the other hand, FI was observed to have minimal effect on AP activation at the levels of PPY988 measured in VH after 36 weeks of treatment.

In vitro modelling of vitreous humor complement activation inhibition by FI vs pegcetacoplan and additional complement regulators. FI (orange), FH (purple), and pegcetacoplan (blue) were titrated (x-axis, ng/mL) through artificial vitreous complement matrix on LPS-coated WeislabTM plates and C3a generation was measured by ELISA (y-axis, ng/mL). C3a is a marker of convertase activity in the complement activation cascade, the potency of complement inhibitors or regulators is reflected in lower amount of C3a generated in the assay when testing comparable concentrations. Mean is shown with +/−SD represented by bars and 4-parameter fit logistic regression curves are shown by lines. The 1st-3rd quartile of increased FI expression from baseline after 36 weeks of PPY988 delivery is shown by orange area. The calculated trough concentration of pegcetacoplan in eyes 27 days after IVT injection is shown by the blue vertical line.

Discussion

Gene therapy has been suggested as a promising treatment approach for slowing macular degeneration in GA secondary to AMD. This manuscript describes a comprehensive examination of the pharmacokinetic / pharmacodynamic properties of complement biomarkers in the context of the PPY988 GA subretinal gene therapy program and provides technical considerations for analyzing biomarker data from treated subjects. Despite the early termination of the clinical development program, the PPY988 study achieved notable milestones in the field, including the implementation of subretinal delivery of gene therapy for GA and the use of serial vitreous sampling. Herein, we present for the first time an analysis of VH complement biomarkers at baseline and after surgical delivery of a subretinal gene therapy. Baseline levels were similar to a previous study of VH complement 35 with potentially confounding loss of total protein also observed in our study, highlighting the need to analyze data in the context of changing total protein levels post-vitrectomy. Our observations reveal consistent elevation of VH FI levels over a 96-week period, accompanied by reduced detection of the FB activation product Ba and ratios of Ba:FB, consistent with sustained modulation of complement AP activity in the vitreous humor compartment. These data confirm that AAV gene therapy has a consistent and lasting effect. Furthermore, although unpowered due to low N, the effects of FI gene therapy on complement breakdown product biomarker profiles (Ba (Figure 2A) and C3b/iC3b (Supplementary Figure 3A)) appear to be comparable in both CFI RV positive and negative sub-populations, which is counter to the original hypothesis of a more dramatic effect in a sub-population with lower baseline FI levels 19,20. However, due to small sample size this finding should be interpreted with caution. Although the impact on C3 breakdown products was minimal at week 36, it became significant at week 96. Elevation of FI and modulation of the broader complement system was also observed in Olink proteomic profiling of VH samples (Supplementary Figure 6), supporting the complement biomarker findings in the validated panel. Elevated levels of FI were detected in both AH and VH. PPY988 efflux from the sub-retinal injection may have transduced cells in the anterior chamber of the eye resulting in an increase in AH FI concentration. However, the correlation of other complement biomarkers between VH and AH was inconsistent, in particular, Ba and C1q exhibited disparate post-treatment trends (Figure 4), indicating that AH alone may not be a reliable biomarker matrix for assessing the activity of complement biomarkers at the target tissues. This cautionary advice should be considered in context of previous reports, suggesting AH could be a broadly reliable proxy for VH 36, albeit that only full length complement proteins were studied therein. To ensure quality control, protein level normalization was added to the analysis comparing biomarker levels before and after vitrectomy. Interestingly, the overall conclusions remained consistent regardless of protein adjustment, which strengthens confidence in the study’s findings. After determining the futility of the Phase II study (HORIZON, NCT04566445), an in vitro assay using clinically relevant levels of complement proteins was conducted to investigate a possible biomarker-related explanation. The results showed that pegcetacoplan had higher potency than FI in reducing C3 convertase activity in vitro suggesting that limited in-vitro potency of FI may have contributed to the lack of clinical success at Phase II. Although the translatability of the VH modelling assay to GA remains to be demonstrated, FI’s lower potency suggests that further increasing FI concentration on its own by gene therapy may not achieve clinically meaningful levels of C3 inhibition. It should however be noted that intravitreal pegcetacoplan will be at its highest concentration in the mid-vitreous whereas FI will most likely be at the highest level in the subretinal space where it is being produced. Hence directly comparing vitreous levels of each compound may underestimate the activity of FI compared with pegcetacoplan at the level of the subretinal space. Notably, that FI does not reduce C3a breakdown to baseline even at supermolecular concentrations suggests cofactor limitation that might be more pronounced in VH given FH is not in excess of FI as is the case in blood 27. Moreover, there are additional cell-bound cofactors for FI that may be present in retinal tissue that are not present in the VH and could further alter the kinetics of the assay, such as MCP (CD46) albeit with disease related changes observed 37. However, the findings of elevated FI in the VH after sub-retinally delivered CFI gene therapy and changes in complement pathway proteins post-treatment build confidence that VH matrix is at least partially reflecting the complement system at the retinal layers and treatment site, and is a valid biomarker for PK/PD insights in response to treatment.

The lack of efficacy in the PPY988 Phase II study does not detract from the viability of complement activation as an effective target for intervention in GA, given the therapeutic effect observed for complement inhibitors pegcetacoplan and avacincaptad pegol (Izervay; Iveric Bio) 5,6, and this study’s observation that pegcetacoplan achieves a much greater potency than FI in vitro. FI had a disproportionally larger effect upon Ba compared to C3a and C3b/iC3b, which may be due to the lability of C3a, differences in diffusion capabilities of the molecules from the retina, and possible contribution of C3 concentration from the choriocapillaris as well as biosynthesis by local immune cells. In addition, we hypothesize FI’s capacity to regulate the production of Ba without causing significant changes in the levels of C3b and C3a in the fluid phase could be contributed to by regulation of AP tickover, resulting from binding and activation of FB on water-hydrolysed C3 (C3H2O) followed by release of Ba10. Similarly, the intravitreal administration of a neutralizing factor D Fab, which was found to be ineffective in Phase III trials for GA, was also able to reduce ocular Ba levels 29. However, it did not have a significant impact on the generation of VH C3 breakdown products. In this study, by week 96, there was an observed rise in the levels of full-length C3 and a corresponding decrease in the ratios of breakdown products to full length. While the specific implications of this change are uncertain, it potentially suggests a reduction in C3 consumption, which aligns with the known activity of FI, but this may be confounded with upregulated biosynthesis of C3 given its acute phase nature 38. C3 breakdown products only partially diffuse due to cell-surface adherence. Based on this understanding, we hypothesize that changes in C3 breakdown products may only become detectable when a certain threshold of modulation in tissue complement activity is reached. Overall, the data suggests that fully functional FI is being secreted into the VH, but the regulatory effects on the level of Ba may be representative of convertase formation in the VH and not the macula retina/RPE nor the choroid. To validate this hypothesis, one approach would be to conduct vitreal sampling using an effective drug targeting C3 for GA in a larger cohort. Further investigations are needed to gain a better understanding of these possibilities.

The study has several limitations that should be considered when interpreting the results. Firstly, the study did not include a control group, which means that the observed changes in complement biomarkers may not be solely attributed to transgene expression. Other factors, such as the impact of subretinal gene therapy administration triggering local inflammation or the effects of total vitrectomy on the protein environment, could also contribute to these changes. During the study, no significant safety concerns or adverse events were reported aside from dose-responsive incidents of RPE changes in the area of the subretinal injection. However, increased levels of antibodies to the vector and transgene were observed following treatment 39. It is important to note that these antibody levels were transient and resolved shortly after treatment, indicating that the observed antibody responses were not persistent or associated with any significant clinical consequences. It is worth noting that the subjects included in this early clinical trial had more advanced disease compared to the typical clinical trial GA population which may impact the generalizability of the findings to the broader GA population. Additionally, the study had a relatively small sample size, which increases the risk of bias. Furthermore, the biomarker conclusions drawn from the study subjects may not be representative of the subjects treated in the subsequent Phase II study. Therefore, caution should be exercised when generalizing the findings to larger populations. Despite convincing and robust genetic links to AMD reported for Type I CFI variants 19–22, this study observed no difference in biomarker response when comparing CFI RV carriers to the broad population. Interestingly, emerging data have revealed more modest ORs for CFI RVs in progression from early to late AMD 23 and macular thinning was observed in UK Biobank eyes of carriers <60 years old 40. Moreover, the long-term effects of complement therapy on the progression of GA are still uncertain. Another limitation to consider is the translatability of the in vitro VH modelling assay to the disease processes occurring in GA, and the relevance of the assay to cell-surface complement activation occurring in the area of GA. The LPS-coated surface used for the VH modelling assay would most likely result in more acute complement activation than in the GA retina. Finally, this is a relative potency assay and does not take into consideration the biodistribution of different modalities and routes of administration. Despite these limitations, the conclusions drawn from this study are compelling and hold significant implications for other gene therapy programs. The introduction of the innovative in vitro complement activation assay serves as a valuable benchmark for gene therapy programs.

The superior potency of FH compared to FI highlights the importance of conducting potency comparisons using in vitro assays that reflect physiologically relevant levels of complement proteins to guide decision-making. This finding opens promising avenues for further investigation, with potential for engineering of more active RCA protein truncates and chimeras including CR1, MCP, DAF, FH, and FHR dimerization domains. To advance gene therapy programs, future research should focus on building upon the pharmacokinetic / pharmacodynamic findings. It is crucial to establish tissue concentrations and estimates of the target required for complement modulation before moving into clinical stages. Pre-clinical assays that can be effectively translated to GA remain elusive. Overall, the role of complement in GA and the potential benefits of a safe one-time gene therapy in reducing AMD progression suggest the need for further clinical development of complement gene therapy.